---
title: Practical Machine Learning Prediction Project
author: "Abdul"
date: "Wednesday, November 19, 2014"
output: html_document
---
             Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Data set).                  
            Data              

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

            What you should submit
The goal of your project is to predict the manner in which they did the exercise. This is the "classes" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

1.Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the write up to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

2.You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

           Reproducible
Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis.             Prepare the data sets
Load the Libraries
---
```{r}
library(caret)
```

```{r}
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```

```{r}
set.seed(12345)
```

Getting the data

Load the training data into a data table.
```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
Load the testing data into a data table.
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```


Read data to memory solely

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

```{r}
getwd()
```


Partitioning the training set into two

Partitioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```
Cleaning the data

The following transformations are used to clean the data:
```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```
code to create another subset without None Zero variables:
```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
#To check the new N?? of observations
dim(myTraining)
```

Killing first column of Data set - ID Removing first ID variable so that it does not interfere with ML Algorithms:
```{r}
myTraining <- myTraining[c(-1)]
```

              Data clean and preprocessing
Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA's I'm going to leave them out:
```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#To check the new N?? of observations
dim(trainingV3)
```


```{r}
myTraining <- trainingV3
rm(trainingV3)
```
Exact the dat for our myTesting and testing data sets.
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#To check the new N?? of observations
dim(myTesting)
```
```{r}
dim(testing)
```
In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.
```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
#And to make sure Coertion really worked, simple smart ass technique:
testing <- rbind(myTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```



Using ML algorithms for prediction: Decision Tree
```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```
The decision tree code
```{r}
fancyRpartPlot(modFitA1)
```

Importance of Variables and Initial Fit of Random Forest Classifier

```{r}
varI <- varImp(modFitA1)
varI <- varI[1:80,]
plot(varI)
```

Prediction models training
The selected prediction models
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```
confusion Matrix to test results
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

```{r}
#Statistics
                                          
#               Accuracy : 0.8683          
#                 95% CI : (0.8607, 0.8757)
#    No Information Rate : 0.2845          
#    P-Value [Acc > NIR] : < 2.2e-16       
                                          
#                  Kappa : 0.8335 
```
Obviously our model performs excellent against the training set, but we need to cross validate the performance against the held out set and see if we have avoided over fitting.

ML algorithms for prediction (Random Forests)
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```
Predicting in-sample error:
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

```{r}
#Statistics
                                         
 #              Accuracy : 0.999          
 #                95% CI : (0.998, 0.9996)
 #   No Information Rate : 0.2845         
 #   P-Value [Acc > NIR] : < 2.2e-16      
                                         
 #                 Kappa : 0.9987         
 #Mcnemar's Test P-Value : NA 
```

Cross-Validation and out of sample error estimating;
With the trained models, can do the cross-validation on the testing data set:

confusion Matrix to test results:
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```
The cross validation accuracy is 99.5% and the out-of-sample error is therefore 0.5% so our model performs rather good.

Error Rate Plot (For All 5 levels)

The out of Bag error rate plot . Black line in the middle is the mse which is the average overall trees
```{r}
par(mar=c(3,4,4,4))                               
plot(modFitB1)                
```



Saving the output for Course Project: Submission

The output files are 20 text files as prediction for each problem_id

```{r}
answers <- predictionsB1
```

```{r}
answers = rep("A", 20)
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(answers)
```

```{r}
answers = rep("B", 20)
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(answers)
```

```{r}
answers = rep("C", 20)
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(answers)
```

```{r}
answers = rep("D", 20)
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(answers)
```

```{r}
answers = rep("E", 20)
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(answers)
```

